# R Data Types and Data Structures

*When human judgement and big data intersect there are some funny things that happen.* 
-Nate Silver 

## Data Types

- As per R's official language definitions; in every computer language variables provide a means of accessing the data stored in memory. 

- R does not provide direct access to the computer's memory but rather provides a number of specialized data structures we will refer to as objects. These objects are referred to through symbols or variables. 

### Double

- Doubles are numbers like 5.0, 5.5, 10.999 etc. They may or may not include decimal places. Doubles are mostly used to represent a continuous variable like serial number, weight, age etc. 

```{r}
x=8.5
is.double(x) #to check if the data type is double
```

### Integer

- Integers are natural numbers. 

```{r}
x=9
typeof(x)
```


```{r}
#The following specifically assigns an integer to x

x=as.integer(9)
typeof(x)
```

### Logical

- A variable of data type logical has the value TRUE or FALSE. To perform calculation on logical objects in R the FALSE is replaced by a zero and TRUE is replaced by 1.

```{r}
x=11
y=10
a=x>y
a
typeof(a)
```



### Character

- Characters represent the string values in R. An object of type character can have alphanumeric strings. Character objects are specified by assigning a string or collection of characters between double quotes (“ string”) . Everything in a double quote is considered a string in R.


